---
title: "Turtles tidied data"
author: "Eric Coyle"
date: "2/21/2021"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(here)
library(tidyverse)
library(naniar)
library(janitor)
library(shiny)
library(paletteer)
library(ggmap)
library(rgeos)
library(rgdal)
library(devtools)
```


```{r}
turtles<-read.csv(here("Final Project","Turtle_data.csv"))
```

```{r}
turtles_tidy<-turtles%>%
  na_if("")%>%
  na_if("0")%>%
  na_if("0.0")%>%
  mutate(dead_alive_new=if_else(Dead_Alive=="alive","Alive",Dead_Alive))
```

```{r}
library(lubridate)
```

```{r}
turtles3<-turtles_tidy%>%
  mutate(DateCapture_new=dmy(DateCapture))%>%
  filter(!is.na(DateCapture_new))%>%
  mutate(Capture_month=month(DateCapture_new))%>%
  mutate(Capture_day=day(DateCapture_new))%>%
  mutate(capture_week_day=wday(DateCapture_new))
head(turtles3)
```

#mapping for release and capture sites

```{r}
turtles3%>%
  select(Cap_Latitude,Cap_Longitude)%>%
  summary()
```

```{r}
cap_lat <- c(33.49, 38.41)
cap_long <- c(-79.08, -75.47)
bbox <- make_bbox(cap_long, cap_lat, f = 0.1)
```

```{r}
cap_map_base <- get_map(bbox, maptype = "terrain-background", source = "stamen")
ggmap(cap_map_base)
```



```{r}
ggmap(cap_map_base) + 
  geom_point(data = turtles3, aes(Cap_Longitude,Cap_Latitude,color=Species,shape=dead_alive_new), size = 2, alpha = 0.7) +
           labs(x = "Longitude", y = "Latitude", title = "Capture Locations")
```
```{r}
turtles3%>%
  select(Rel_Latitude,Rel_Longitude)%>%
  summary()
```

```{r}
rel_lat <- c(27.82, 39)
rel_long <- c(-81.24, -36.03)
bbox2 <- make_bbox(rel_long, rel_lat, f = 0.05)
```

```{r}
rel_map_base <- get_map(bbox2, maptype = "terrain-background", source = "stamen")
ggmap(rel_map_base)
```
```{r}
ggmap(rel_map_base) + 
  geom_point(data = turtles3, aes(Rel_Longitude,Rel_Latitude,color=Species), size = 2, alpha = 0.7) +
           labs(x = "Longitude", y = "Latitude", title = "Release Locations")
```
#we can then make additional maps with color or shape of the dots for different variables. We also should pick a unifying color scheme/text/chart and map theme for all of our images (Ideally a color that stands out against the terrain background if we want to stick with terrain for our maps)

#The following are some ideas for graphs that I thought would be good to include, but we can each come up and explore some of our own to compile into the final product

```{r}
turtles3%>%
  group_by(year)%>%
  summarise(n())
```
```{r}
turtles3%>%
  group_by(year)%>%
  ggplot(aes(x=year,fill=dead_alive_new))+
  geom_bar(position = "dodge")
```


```{r}
turtles3%>%
  group_by(Capture_month)%>%
  summarise(n())
```
```{r}
turtles3%>%
  ggplot(aes(month(x=DateCapture_new,label=TRUE)))+
  geom_bar(position = "dodge")+
  labs(x = NULL,
         y = "Number of Captures")
```


```{r}
turtles3%>%
  group_by(Capture_day)%>%
  summarise(n())
```

```{r}
turtles3%>%
  ggplot(aes(day(x=DateCapture_new)))+
  geom_bar()+
  labs(x = NULL,
         y = "Number of Captures")
```


```{r}
turtles3%>%
  group_by(capture_week_day)%>%
  summarise(n())
```
```{r}
turtles3%>%
  ggplot(aes(wday(x=DateCapture_new,label=TRUE)))+
  geom_bar()+
  labs(x = NULL,
         y = "Number of Captures")
```
#Monday and Friday are popular trapping days apparently

```{r}
turtles3%>%
  group_by(year)%>%
  mutate(mean_body_length=mean(SCL_tip,na.rm=T))%>%
  ggplot(aes(x=year,y=mean_body_length,group=1,color=Species))+
  geom_line()+
  geom_point()+
  facet_wrap(~Species)
```
```{r}
turtles3%>%
  ggplot(aes(x=Species,y=CCL_tip,color=Species))+
  geom_boxplot()
```
#we of course can fix the axis and replicate this kind of graph for other measurements (SCL_noth/SCl_tip, CCl_notch, etc)

```{r}
turtles3%>%
  filter(Gear!=""&Species!="")%>%
  ggplot(aes(x=Gear,fill=Species))+
  geom_bar(position = "dodge")+
  labs(title = "Gear used for Different Species",x="Gear",y="Number Captured with Given Method")+
  theme(axis.text.x = element_text(angle = 60,hjust = 1))+
  scale_y_log10()
```

