---
title: "Midterm 2"
author: "Eric Coyle"
date: "`r Sys.Date()`"
output:
  html_document: 
    theme: spacelab
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Instructions
Answer the following questions and complete the exercises in RMarkdown. Please embed all of your code and push your final work to your repository. Your code should be organized, clean, and run free from errors. Be sure to **add your name** to the author header above. You may use any resources to answer these questions (including each other), but you may not post questions to Open Stacks or external help sites. There are 10 total questions.  

Make sure to use the formatting conventions of RMarkdown to make your report neat and clean! Your plots should use consistent aesthetics throughout. Feel free to be creative- there are many possible solutions to these questions!  

This exam is due by **12:00p on Tuesday, February 23**.  

## Load the libraries
```{r}
library(tidyverse)
library(janitor)
library(here)
options(scipen=999) #disables scientific notation when printing
```

## Gapminder
For this assignment, we are going to use data from  [gapminder](https://www.gapminder.org/). Gapminder includes information about economics, population, social issues, and life expectancy from countries all over the world. We will use three data sets, so please load all three.  

One thing to note is that the data include years beyond 2021. These are projections based on modeling done by the gapminder organization. Start by importing the data.
```{r}
population <- read_csv(here("midterm2","data", "population_total.csv"))
population
```

```{r}
income <- read_csv(here("midterm2","data", "income_per_person_gdppercapita_ppp_inflation_adjusted.csv"))
income
```

```{r}
life_expectancy <- read_csv(here("midterm2","data", "life_expectancy_years.csv"))
life_expectancy
```

```{r}
#glimpse(population)
#glimpse(income)
#glimpse(life_expectancy)
```

1. (3 points) Once you have an idea of the structure of the data, please make each data set tidy and store them as new objects. You will need both the original and tidy data!
```{r}
population_tidy<-population%>%
  pivot_longer(-country,names_to = "year",values_to = "population")
population_tidy
```

```{r}
income_tidy<-income%>%
  pivot_longer(-country,names_to = "year",values_to = "income")
income_tidy
```

```{r}
life_expectancy_tidy<-life_expectancy%>%
  pivot_longer(-country,names_to = "year",values_to = "life_expectancy")
life_expectancy_tidy
```
```{r}
summary(income_tidy)
summary(life_expectancy_tidy)
summary(population_tidy)
```
```{r}
life_expectancy_tidy%>%
  arrange(`life_expectancy`)
```

2. (1 point) How many different countries are represented in the data? Provide the total number and their names. Since each data set includes different numbers of countries, you will need to do this for each one.
```{r}
population_tidy%>%
  group_by(country)%>%
  summarise(total_years_of_data=n())
```
#195 different countries in the population data
```{r}
income_tidy%>%
  group_by(country)%>%
  summarise(total_years_of_data=n())
```
#193 countries in the income data
```{r}
life_expectancy_tidy%>%
  group_by(country)%>%
  summarise(total_years_of_data=n())
```
#187 countries in the life expectancy data

## Life Expectancy  

3. (2 points) Let's limit the data to the past 100 years (1920-2020). For these years, which country has the highest life expectancy? How about the lowest life expectancy?  

Highest
```{r}
life_expectancy_tidy%>%
  group_by(country)%>%
  filter(year<=2020&year>=1920)%>%
  arrange(desc(life_expectancy))
```
#highest individual life expectacy during this span was in Singapore in 2020 at 85.3 years

```{r}
life_expectancy_tidy%>%
  group_by(country)%>%
  filter(year<=2020&year>=1920)%>%
  summarise(mean_life_expectancy=mean(life_expectancy,na.rm=T))%>%
  arrange(desc(mean_life_expectancy))
```
#highest mean life expectancy over these 100 years was Andorra
```{r}
life_expectancy_tidy%>%
  filter(country=="Andorra")
```
#Andorra is highest when NA's are removed becuase it is a recent country (1970 is when data begins), so if NAs are not removed the highest life expectancy over the 100 year span is in Sweden
#I decided to leave NA's in to not bias data

Lowest
```{r}
life_expectancy_tidy%>%
  group_by(country)%>%
  filter(year<=2020&year>=1920)%>%
  arrange(life_expectancy)
```
```{r}
life_expectancy_tidy%>%
  group_by(country)%>%
  filter(year<=2020&year>=1920)%>%
  summarise(mean_life_expectancy=mean(life_expectancy,na.rm=T))%>%
  arrange((mean_life_expectancy))
```
#lowest single year life expectancy was in Kazakhstan in 1933- 4.07 years! and Lowest average life expectancy over the 100 year period was the Central African Republic with 41.76634 years

4. (3 points) Although we can see which country has the highest life expectancy for the past 100 years, we don't know which countries have changed the most. What are the top 5 countries that have experienced the biggest improvement in life expectancy between 1920-2020?
```{r}
life_expectancy_tidy%>%
  group_by(country)%>%
  filter(year==2020|year==1920)%>%
  mutate(Life_expectancy_growth = life_expectancy- lag(life_expectancy, default = first(life_expectancy)))%>%
  arrange(desc(Life_expectancy_growth))
```
#top five are Kuwait, Kyrgyz Republic, Turkmenistan, South Korea, and Tajikistan

```{r}
life_expectancy_tidy%>%
  group_by(country)%>%
  filter(year==2020|year==1920)%>%
  mutate(Life_expectancy_growth = life_expectancy- lag(life_expectancy, default = first(life_expectancy)))%>%
  arrange(desc(Life_expectancy_growth))%>%
  filter(Life_expectancy_growth>54.2)
```

```{r}
life_expectancy_tidy%>%
  group_by(country)%>%
  filter(year==2020|year==1920)%>%
  mutate(Life_expectancy_growth = life_expectancy- lag(life_expectancy, default = first(life_expectancy)))%>%
  arrange(desc(Life_expectancy_growth))%>%
  filter(Life_expectancy_growth>54.2)%>%
  ggplot(aes(x=country,y=Life_expectancy_growth,fill=country))+
  geom_col()+
  scale_fill_brewer(palette = "Set1")+
  theme(axis.text.x = element_text(angle = 60, hjust = 1))+
  labs(title = "Five countries with the most improvement in Life Expectancy from 1920-2020",x="Country",y="Improvement in Life Expectancy")
```
<style>
div.blue { background-color:#e6f0ff; border-radius: 5px; padding: 20px;}
</style>
<div class = "blue">

5. (3 points) Make a plot that shows the change over the past 100 years for the country with the biggest improvement in life expectancy. Be sure to add appropriate aesthetics to make the plot clean and clear. Once you have made the plot, do a little internet searching and see if you can discover what historical event may have contributed to this remarkable change.  
```{r}
life_expectancy_tidy_1920_2020<-life_expectancy_tidy%>%
  filter(year<=2020&year>=1920)
```


```{r}
life_expectancy_tidy_1920_2020$year<-as.factor(life_expectancy_tidy_1920_2020$year)
```

```{r}
life_expectancy_tidy_1920_2020%>%
  filter(country=="Kuwait")%>%
  ggplot(aes(x=year,y=life_expectancy,group=1))+
  geom_line()+
  geom_point(shape=2)+
  scale_fill_brewer(palette = "Set1")+
  theme(axis.text.x = element_text(angle = 60, hjust = 1))+
  scale_x_discrete(breaks=c(1920,1940,1960,1980,2000,2020))+
  labs(title = "Life Expectancy from 1920-2020 in Kuwait",x="Year",y="Life Expectancy")
```
</div>


## Population Growth
6. (3 points) Which 5 countries have had the highest population growth over the past 100 years (1920-2020)?
```{r}
population_tidy%>%
  group_by(country)%>%
  filter(year==2020|year==1920)%>%
  mutate(population_growth = population- lag(population, default = first(population)))%>%
  arrange(desc(population_growth))
```
```{r}
population_tidy%>%
  group_by(country)%>%
  filter(year==2020|year==1920)%>%
  mutate(population_growth = population- lag(population, default = first(population)))%>%
  arrange(desc(population_growth))%>%
  filter(population_growth>185400000)
```

7. (4 points) Produce a plot that shows the 5 countries that have had the highest population growth over the past 100 years (1920-2020). Which countries appear to have had exponential growth?  
```{r}
population_tidy_1920_2020<-population_tidy%>%
  filter(year<=2020&year>=1920)
```

```{r}
population_tidy_1920_2020$year<-as.factor(population_tidy_1920_2020$year)
```


```{r}
population_tidy_1920_2020%>%
  filter(country=="China"|country=="India"|country=="Indonesia"|country=="United States"|country=="Pakistan")%>%
  ggplot(aes(x=year,y=population,group=country,color=country))+
  geom_line()+
  geom_point(shape=1)+
  scale_fill_brewer(palette = "Set1")+
  theme(axis.text.x = element_text(angle = 60, hjust = 1))+
  scale_x_discrete(breaks=c(1920,1940,1960,1980,2000,2020))+
  labs(title = "Five countries with the Highest Population Growth from 1920-2020",x="Year",y="Life Expectancy")
```
#China and India have aproximately exponential growth for the 1920-2020 time frame

## Income
The units used for income are gross domestic product per person adjusted for differences in purchasing power in international dollars.

<style>
div.blue { background-color:#e6f0ff; border-radius: 5px; padding: 20px;}
</style>
<div class = "blue">

8. (4 points) As in the previous questions, which countries have experienced the biggest growth in per person GDP. Show this as a table and then plot the changes for the top 5 countries. With a bit of research, you should be able to explain the dramatic downturns of the wealthiest economies that occurred during the 1980's.
```{r}
income_tidy%>%
  group_by(country)%>%
  filter(year==2020|year==1920)%>%
  mutate(GDP_per_capita_growth = income- lag(income, default = first(income)))%>%
  arrange(desc(GDP_per_capita_growth))
```
```{r}
income_tidy%>%
  group_by(country)%>%
  filter(year==2020|year==1920)%>%
  mutate(GDP_per_capita_growth = income- lag(income, default = first(income)))%>%
  filter(GDP_per_capita_growth>63210)%>%
  arrange(desc(GDP_per_capita_growth))
```
```{r}
income_tidy_1920_2020<-income_tidy%>%
  filter(year<=2020&year>=1920)
```

```{r}
income_tidy_1920_2020%>%
  filter(country=="Qatar"|country=="Luxembourg"|country=="Singapore"|country=="Brunei"|country=="Ireland")%>%
  ggplot(aes(x=year,y=income,group=country,color=country))+
  geom_line()+
  geom_point(shape=1)+
  scale_fill_brewer(palette = "Set1")+
  theme(axis.text.x = element_text(angle = 60, hjust = 1))+
  scale_x_discrete(breaks=c(1920,1940,1960,1980,2000,2020))+
  labs(title = "Five countries with the Highest Growth in GDP/Capita 1920-2020",x="Year",y="Life Expectancy")
```
</div>

9. (3 points) Create three new objects that restrict each data set (life expectancy, population, income) to the years 1920-2020. Hint: I suggest doing this with the long form of your data. Once this is done, merge all three data sets using the code I provide below. You may need to adjust the code depending on how you have named your objects. I called mine `life_expectancy_100`, `population_100`, and `income_100`. For some of you, learning these `joins` will be important for your project.  

life_expectancy_100
```{r}
Life_expectancy_tidy_1920_2020<-life_expectancy_tidy%>%
  filter(year<=2020&year>=1920)
```

population_100
```{r}
population_tidy_1920_2020<-population_tidy%>%
  filter(year<=2020&year>=1920)
```

income_tidy_1920_2020
```{r}
income_tidy_1920_2020<-income_tidy%>%
  filter(year<=2020&year>=1920)
```

```{r}
gapminder_join <- inner_join(life_expectancy_tidy_1920_2020, population_tidy_1920_2020, by= c("country", "year"))
gapminder_join <- inner_join(gapminder_join, income_tidy_1920_2020, by= c("country", "year"))
gapminder_join
```

10. (4 points) Use the joined data to perform an analysis of your choice. The analysis should include a comparison between two or more of the variables `life_expectancy`, `population`, or `income.`
```{r}
gapminder_join%>%
  group_by(country)%>%
  mutate(maxlife=max(life_expectancy),minlife=min(life_expectancy))%>%
  filter(life_expectancy==maxlife|life_expectancy==minlife)
```


```{r}
gapminder_join%>%
  group_by(country)%>%
  mutate(maxlife=max(life_expectancy),minlife=min(life_expectancy))%>%
  mutate(max_difference=maxlife-minlife)%>%
  filter(life_expectancy==maxlife|life_expectancy==minlife)%>%
  filter(country=="Qatar"|country=="Luxembourg"|country=="Singapore"|country=="Brunei"|country=="Ireland")
```
```{r}
gapminder_join%>%
  group_by(country)%>%
  mutate(maxlife=max(life_expectancy),minlife=min(life_expectancy))%>%
  mutate(max_difference=maxlife-minlife)%>%
  filter(life_expectancy==maxlife|life_expectancy==minlife)%>%
  filter(country=="Qatar"|country=="Luxembourg"|country=="Singapore"|country=="Brunei"|country=="Ireland")%>%
  ggplot(aes(x=year,y=maxlife,fill=country))+
  geom_col(position = "dodge")+
  scale_fill_brewer(palette = "Set1")+
  theme(axis.text.x = element_text(angle = 60, hjust = 1))+
  labs(title = "Years With Max Life Expectancy in Five countries with Highest GDP 1920-2020",x="Year",y="Length of Maximum Life Expectancy")
gapminder_join%>%
  group_by(country)%>%
  mutate(maxlife=max(life_expectancy),minlife=min(life_expectancy))%>%
  mutate(max_difference=maxlife-minlife)%>%
  filter(life_expectancy==maxlife|life_expectancy==minlife)%>%
  filter(country=="Qatar"|country=="Luxembourg"|country=="Singapore"|country=="Brunei"|country=="Ireland")%>%
  ggplot(aes(x=year,y=minlife,fill=country))+
  geom_col(position = "dodge")+
  scale_fill_brewer(palette = "Set1")+
  theme(axis.text.x = element_text(angle = 60, hjust = 1))+
  labs(title = "Years With Min Life Expectancy in Five countries with Highest GDP 1920-2020",x="Year",y="Length of Shortest Life expectancy")
```

