---
title: "Lab 9 Homework"
author: "Eric Coyle"
date: "`r Sys.Date()`"
output:
  html_document: 
    theme: spacelab
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Instructions
Answer the following questions and complete the exercises in RMarkdown. Please embed all of your code and push your final work to your repository. Your final lab report should be organized, clean, and run free from errors. Remember, you must remove the `#` for the included code chunks to run. Be sure to add your name to the author header above.  

Make sure to use the formatting conventions of RMarkdown to make your report neat and clean!  

## Load the libraries
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
library(here)
library(naniar)
``` 

For this homework, we will take a departure from biological data and use data about California colleges. These data are a subset of the national college scorecard (https://collegescorecard.ed.gov/data/). Load the `ca_college_data.csv` as a new object called `colleges`.
```{r}
colleges<-read.csv(here("lab9","data","ca_college_data.csv"))
colleges
```

The variables are a bit hard to decipher, here is a key:  

INSTNM: Institution name  
CITY: California city  
STABBR: Location state  
ZIP: Zip code  
ADM_RATE: Admission rate  
SAT_AVG: SAT average score  
PCIP26: Percentage of degrees awarded in Biological And Biomedical Sciences  
COSTT4_A: Annual cost of attendance  
C150_4_POOLED: 4-year completion rate  
PFTFTUG1_EF: Percentage of undergraduate students who are first-time, full-time degree/certificate-seeking undergraduate students  

1. Use your preferred function(s) to have a look at the data and get an idea of its structure. Make sure you summarize NA's and determine whether or not the data are tidy. You may also consider dealing with any naming issues.
```{r}
glimpse(colleges)
```

```{r}
colleges%>%
  naniar::miss_var_summary()
```

```{r}
colleges2<-colleges%>%
  janitor::clean_names()
colleges2
```
```{r}
colleges2%>%
  naniar::miss_var_summary()
head(colleges2)
```

2. Which cities in California have the highest number of colleges?
```{r}
college_counts<-colleges2%>%
  count(city)%>%
  arrange(desc((n)))
college_counts
```
#pretty quick drop off

3. Based on your answer to #2, make a plot that shows the number of colleges in the top 10 cities.
```{r}
college_counts%>%
  top_n(10,n)%>%
  ggplot(aes(x=reorder(city,n),y=n))+
  geom_col()+
  coord_flip()
```

4. The column `COSTT4_A` is the annual cost of each institution. Which city has the highest average cost? Where is it located?
```{r}
colleges2%>%
  group_by(city)%>%
  summarise(ave_cost=mean(costt4_a,na.rm=T))%>%
  arrange(desc(ave_cost))
```
#those fancy claremont colleges...

5. Based on your answer to #4, make a plot that compares the cost of the individual colleges in the most expensive city. Bonus! Add UC Davis here to see how it compares :>).
```{r}
colleges2%>%
  filter(city=="Davis")
```
```{r}
colleges2%>%
  filter(city=="Claremont"|city=="Davis")%>%
  group_by(costt4_a)
```


```{r}
colleges2%>%
  filter(city=="Claremont"|city=="Davis")%>%
  group_by(costt4_a)%>%
  ggplot(aes(x=reorder(instnm,costt4_a),y=costt4_a))+
  geom_col()+
  coord_flip()+
  labs(title = "Davis_v_Claremont")
```
#Davis is a bargain (relatively speaking of course)!

6. The column `ADM_RATE` is the admissions rate by college and `C150_4_POOLED` is the four-year completion rate. Use a scatterplot to show the relationship between these two variables. What do you think this means?
```{r}
colleges2%>%
  ggplot(aes(x=adm_rate,y=c150_4_pooled))+
  geom_jitter()+
  geom_smooth(method = lm,se=T)
```
#the more students admitted, the lower the completion rate, which makes sense as more students means more people who might drop out. Also, one might expect those attending and accepted by a more selective college would want to go there and finish their degree since it was hard to get accepted in the first place.

7. Is there a relationship between cost and four-year completion rate? (You don't need to do the stats, just produce a plot). What do you think this means?
```{r}
colleges2%>%
  ggplot(aes(x=costt4_a,y=c150_4_pooled))+
  geom_jitter()
```
#there appears to be a moderate to weak positive correlation between cost and completion at colleges, possibly reflecting increased motivation to finish a degree if one has already paid so much to attend for a given ampount of time, but there are many other variables and nuances at play as well.

8. The column titled `INSTNM` is the institution name. We are only interested in the University of California colleges. Make a new data frame that is restricted to UC institutions. You can remove `Hastings College of Law` and `UC San Francisco` as we are only interested in undergraduate institutions.
```{r}
UC_system<-colleges2%>%
  filter(str_detect(instnm,"University of California"))
UC_system
```

Remove `Hastings College of Law` and `UC San Francisco` and store the final data frame as a new object `univ_calif_final`.
```{r}
univ_calif_final<-UC_system%>%
  top_n(8,sat_avg)
univ_calif_final
```
#those other two have no dat associated with them so this worked in thsie case but will f course not apply to all data sets
Use `separate()` to separate institution name into two new columns "UNIV" and "CAMPUS".
```{r}
univ_calif_final2<-univ_calif_final%>%
  separate("instnm",into = c("univ","campus"),"-")
univ_calif_final2
```

9. The column `ADM_RATE` is the admissions rate by campus. Which UC has the lowest and highest admissions rates? Produce a numerical summary and an appropriate plot.
```{r}
univ_calif_final2%>%
  group_by(adm_rate)%>%
  select(univ,campus,adm_rate)%>%
  arrange(desc(adm_rate))
```

```{r}
univ_calif_final2%>%
  ggplot(aes(x=campus,y=adm_rate))+
  geom_col()
```

10. If you wanted to get a degree in biological or biomedical sciences, which campus confers the majority of these degrees? Produce a numerical summary and an appropriate plot.
```{r}
names(colleges2)
```

```{r}
colleges2%>%
  top_n(10,pcip26)%>%
  select(instnm,pcip26)%>%
  arrange(desc(pcip26))
```

```{r}
colleges2%>%
  top_n(10,pcip26)%>%
  ggplot(aes(x=reorder(instnm,pcip26),y=pcip26))+
  geom_col()+
  coord_flip()
```

## Knit Your Output and Post to [GitHub](https://github.com/FRS417-DataScienceBiologists)